The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-10-18 talk/vmil2012/presentation/talk.tex
use same color for group and single plot
interleave example with explanation slides

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \begin{itemize}
          \item for interesting, commonly executed code paths
          \item produces a linear list of operations (trace)
          \item automatically does (potentially deep) inlining
      \end{itemize}
      \item trace is optimized and then turned into machine code
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Guards}
  \begin{itemize}
      \item Points of control flow divergence are marked with guards
      \item Operations that check whether conditions are still true
      \item When a guard fails, execution of the trace stops and continues in the interpreter
      \pause
      \item \emph{This talk:} describe technology and design decisions around guards
      \pause
      \begin{block}{Guard Characteristics}
          \begin{itemize}
              \item lots of them, up to 20\% guards
              \item most never fail
              \item need big information attached to them
          \end{itemize}
      \end{block}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Running Example}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Interaction with Optimization}
  \begin{itemize}
      \item Some optimizations make it necessary to store extra information in symbolic frames
      \item examples:
          \begin{itemize}
              \item allocation removal (need to allocate objects before resuming)
              \item delayed heap stores (need to do stores before resuming interpreter)
          \end{itemize}
      \item can be compressed using similar techniques
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Emitting Guards}
  Guards are compiled as
  \begin{itemize}
    \item Quick Check if the condition holds
    \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
  \end{itemize}
  In case of failure
  \begin{itemize}
    \item execution jumps to shared compensation code, decodes and stores mapping
    \item returns to interpreter that rebuilds state
  \end{itemize}
\end{frame}

\begin{frame}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/loop.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
  \end{figure}
\end{frame}
\begin{frame}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Patching Guards for Bridges}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates / Go Benchmark}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/go_data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Tracing JITs Compile by Observing an Interpreter}
  \begin{itemize}
      \item VM contains both an interpreter and the tracing JIT compiler
      \item JIT works by observing and logging what the interpreter does
      \item for interesting, commonly executed code paths
      \item produces a linear list of operations (trace)
  \end{itemize}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop01}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop02}
\end{frame}

\begin{frame}
  \frametitle{Guards}
  \begin{itemize}
      \item Points of control flow divergence are marked with guards
      \item Operations that check whether conditions are still true
      \item When a guard fails, execution of the trace stops and continues in the interpreter
      \pause
      \item \emph{This talk:} technology and design decisions of guards
      \pause
      \begin{block}{Guard Characteristics}
          \begin{itemize}
              \item lots of them, up to 20\% guards
              \item most never fail
              \item need big information attached to them
          \end{itemize}
      \end{block}
  \end{itemize}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop03}
\end{frame}

\begin{frame}
  \frametitle{Inlining}
  Tracing automatically does (potentially deep) inlining
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop04}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop05}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop06}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop07}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop08}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop09}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop10}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop11}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop12}
\end{frame}

@@

@@

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop07}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/framechain1}
\end{frame}

@@

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop07}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/framechain1}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/loop08}
\end{frame}

\begin{frame}
  \includegraphics[scale=0.4]{figures/framechain2}
\end{frame}

\begin{frame}
  \frametitle{Interaction with Optimization}
  \begin{itemize}
      \item Some optimizations make it necessary to store extra information in symbolic frames
      \pause
      \item examples:
          \begin{itemize}
              \item allocation removal (need to allocate objects before resuming)
              \item delayed heap stores (need to do stores before resuming interpreter)
          \end{itemize}
      \item can be compressed using similar techniques
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Emitting Guards}
  Guards are compiled as
  \begin{itemize}
    \item quick Check if the condition holds
    \item and a mapping of machine locations to JIT-variables % indirection using the fail-boxes
  \end{itemize}
  \pause
  In case of failure
  \begin{itemize}
    \item execution jumps to shared compensation code, decodes and stores mapping
    \item returns to interpreter that rebuilds state
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Compiling a Trace}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/loop.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Compiling a Bridge}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_compiled.pdf}
  \end{figure}
\end{frame}
\begin{frame}
  \frametitle{Patching Guards for Bridges}
  \begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{figures/bridge_patched.pdf}
  \end{figure}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}
\begin{frame}[t,fragile]
\pgfplotsset{tick label style={font=\tiny\bfseries},
label style={font=\small},
legend style={font=\tiny}
}
    \frametitle{Guard Failure Rates / Go Benchmark}
    \begin{figure}
        \centering
        \begin{tikzpicture}
            \begin{axis}[
            xlabel= Guards by failures,
            ylabel=Relative \# of failures,
            xtick=\empty,
            ytick=\empty,
            ]
                \input{figures/go_data}
            \end{axis}
        \end{tikzpicture}
    \end{figure}
\end{frame}
